An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-03-30 EconGrowth-R-Functions.Rnw
Now possible to include the optimum point on the SSE vs. lambda and gamma graph.

#########################
# Creates a 3-D wireframe graph with SSE on the vertical axis and 
# lambda and m on the horizontal axes.
# params: 
#   countryAbbrev: 2-letter country code (upper case)
#   factor: the factor of production you want to use in the single-factor model.
##
sf3DSSEGraph <- function(countryAbbrev, factor){
  # Generate the data in the form of a matrix with gamma in columns, lambda in rows, and SSE in the middle
  # First, make a data.frame with which we want to make predictions
  lambdaSeq <- seq(0.0, 0.05, by=0.01)
  mSeq <- seq(0.0, 1.0, by=0.1)

  lambdaSeq <- data.frame(lambdaSeq)
  nRowLambda <- nrow(lambdaSeq)
  colnames(lambdaSeq) <- "lambda"

  mSeq <- data.frame(mSeq)
  nRowM <- nrow(mSeq)
  colnames(mSeq) <- "m"

  # Make a sequence of lambda values based on the number of m values you have.
  lambda <- rep(lambdaSeq[1,1], nRowM)
  lambda <- data.frame(lambda)
  colnames(lambda) <- "lambda"
  for (i in 2:nRowLambda){
    lambdaTemp <- rep(lambdaSeq[i, 1], nRowM)
    lambdaTemp <- data.frame(lambdaTemp)
    colnames(lambdaTemp) <- "lambda"
    lambda <- rbind(lambda, lambdaTemp)
  }

  # Make repeats of the m data frame as we have number of lambda rows
  m <- mSeq
  for (i in 2:nRowLambda){
    m <- rbind(m, mSeq)
  }

  # Now create the data.frame of new observation locations
  newX <- cbind(lambda, m)

  # Get the actual data
  data <- loadData(countryAbbrev)
  y_act <- data[ ,"iGDP"] # Pick off the GDP column
  y_act <- data.frame(y_act)

  # Make predictions
  # Get the model
  model <- singleFactorModel(countryAbbrev, factor)

  sse <- c(1:(nRowLambda*nRowM)) # Build a bogus vector at the start

  # Loop over all the different combinations of lambda and m that are of interest for us.
  #i = 1
  for (i in 1:(nRowLambda*nRowM)){
    y_pred <- predict(model, newX[i,]) # Get lambda and m for this particular row.
    y_pred <- data.frame(y_pred)

    compareDF <- cbind(y_pred, y_act)
    se <- with(compareDF, (y_pred-y_act)^2) # Makes a column of squared errors at each year
    compareDF <- cbind(compareDF, se) # Bind it to the data frame as an additional column

    # Now get the sse
    sums <- colSums(compareDF)
    sse[i] <- sums[3] # squared error (se) is the 3rd column
  }
  ssedf <- data.frame(sse)
  colnames(ssedf) <- "sse"

  # Now merge the sse column and the newX data to form a data.frame fit for 3-D plotting
  dataFor3DGraph <- cbind(newX, ssedf)

  # Adjust settings for a nicer graph
  #trellis.par.set("axis.line", list(col="transparent"))

  # Create the lattice graph
  fig <- wireframe(sse ~ lambda * m,
                   data = dataFor3DGraph,
                   screen = list(z = -20, x = -55),
                   xlab = "$\\lambda$",
                   ylab = "$\\gamma$",
                   zlab = list(label="$SSE_y$", rot=90),
                   scales = list(arrows=FALSE, 
                                 cex=0.45, 
                                 col="black", 
                                 tck=0.8,
                                 x=list(at=c(0.00, 0.01, 0.02, 0.03)),
                                 y=list(at=c(0.0, 0.5, 1.0)),
                                 z=list(at=c(10, 20, 30))
                                 ),
                   xlim = c(0.0, 0.04), # lambda axis
                   ylim = c(0.0, 1.0), # gamma axis
                   zlim = c(0, 30) # sse axis
                   )
  return(fig)    
}
@

>>>>

###############################
# Creates a panel for use with a mixed wireframe and cloud.
# See http://stackoverflow.com/questions/1406202/plotting-a-wireframe-and-a-cloud-with-lattice-in-r
# 
# x: x variable to be used in the wireframe plot. This will be the x variable from the wireframe plot.
# y: y variable to be used in the wireframe plot. This will be the y variable from the wireframe plot.
# z: z variable to be used in the wireframe plot. This will be the z variable from the wireframe plot.
# points: a data.frame containing columns named x, y, and z.
# showPoints: a boolean saying whether you want the points to be shown.
##
wireCloudPanel <- function(x, y, z, points, showPoints=TRUE, ...) {
  panel.wireframe(x, y, z, ...)
  if (showPoints){
    panel.cloud(points$x, points$y, points$z, ...)
  }
}

#########################
# Creates a 3-D wireframe graph with SSE on the vertical axis and 
# lambda and m on the horizontal axes.
# params: 
#   countryAbbrev: 2-letter country code (upper case)
#   factor: the factor of production you want to use in the single-factor model. One of K, L, Q, X, or U
#   showOpt: tells whether or not to show a dot at the optimium point.
##
sf3DSSEGraph <- function(countryAbbrev, factor, showOpt=TRUE){
  # First, make a data.frame with which we want to make predictions
  lambdaSeq <- seq(0.0, 0.05, by=0.01)
  mSeq <- seq(0.0, 1.0, by=0.1)
  newX <- data.frame(expand.grid(lambdaSeq, mSeq))
  colnames(newX) <- c("lambda", "m")
  # Get the actual GDP data
  data <- loadData(countryAbbrev)
  y_act <- data[ ,"iGDP"] # Pick off the GDP column
  y_act <- data.frame(y_act)
  # Need the model to make predictions
  model <- singleFactorModel(countryAbbrev, factor)
  # Get information about the optimum point, the point where SSE is minimized
  coefs <- coef(model)
  lambda_opt <- coefs["lambda"]
  gamma_opt <- coefs["m"]
  xOpt <- data.frame(lambda=lambda_opt, m=gamma_opt)
  y_opt_pred <- predict(model, xOpt)
  y_opt_pred <- data.frame(y_opt_pred)
  compareDF_opt <- cbind(y_opt_pred, y_act)
  se_opt <- with(compareDF_opt, (y_opt_pred-y_act)^2)
  compareDF_opt <- cbind(compareDF_opt, se_opt)
  sums_opt <- colSums(compareDF_opt)
  sse_opt <- sums_opt[3]
  # Need to name these x, y, and z so that the custom panel will pick them up.
  optPoint <- data.frame(x=lambda_opt, y=gamma_opt, z=sse_opt)
  rownames(optPoint) <- NULL

  # Now gather all the points we need for the surface.
  nObs <- nrow(newX)
  sse <- c(1:nObs) # Build a bogus vector at the start
  # Loop over all the different combinations of lambda and m that are of interest for us.
  for (i in 1:nObs){
    y_pred <- predict(model, newX[i,]) # Get lambda and m for this particular row.
    y_pred <- data.frame(y_pred)
    compareDF <- cbind(y_pred, y_act)
    se <- with(compareDF, (y_pred-y_act)^2) # Makes a column of squared errors at each year
    compareDF <- cbind(compareDF, se) # Bind it to the data frame as an additional column
    # Now get the sse
    sums <- colSums(compareDF)
    sse[i] <- sums[3] # squared error (se) is the 3rd column
  }
  ssedf <- data.frame(sse)
  colnames(ssedf) <- "sse"
  # Now merge the sse column and the newX data to form a data.frame fit for 3-D plotting
  dataFor3DSurface <- cbind(newX, ssedf)
  # Create the figure
  fig <- wireframe(sse ~ lambda * m,
                   data = dataFor3DSurface,
                   panel = wireCloudPanel,
                   points = optPoint,
                   showPoints = showOpt,
                   screen = list(z = -20, x = -55),
                   xlab = "$\\lambda$",
                   ylab = "$\\gamma$",
                   zlab = list(label="$SSE_y$", rot=90),
                   scales = list(arrows=FALSE, 
                                 cex=0.45, 
                                 col="black", 
                                 tck=0.8,
                                 x=list(at=c(0.00, 0.01, 0.02, 0.03)),
                                 y=list(at=c(0.0, 0.5, 1.0)),
                                 z=list(at=c(10, 20, 30))
                                 ),
                   xlim = c(0.0, 0.04), # lambda axis
                   ylim = c(0.0, 1.0), # gamma axis
                   zlim = c(0, 30) # sse axis
                   )
  return(fig)    
}
@

### 2013-03-30 EconGrowth.Rnw
Now including tables and frames for sources of economic data. Also including graphics for world bank and BLS.
Now possible to include the optimum point on the SSE vs. lambda and gamma graph.

<<test3dPlot, eval=TRUE>>=
sf3DSSEGraph("US", "Q")
@

>>>>

<<TestWireframeCloud, eval=TRUE>>=

sf3DSSEGraph("US", "Q", showOpt=TRUE)

# point <- data.frame(x=0.0, y=0.0, z=1.0)
# 
# x <- seq(-.8, .8, .1)
# y <- seq(-.8, .8, .1)
# myGrid <- data.frame(expand.grid(x,y))
# colnames(myGrid) <- c("x","y")
# myGrid$z <- myGrid$x + myGrid$y
# noise <- rnorm(length(myGrid$z),.3,.2)
# myGrid$z2 <- myGrid$x + myGrid$y + noise
# mypanel <- function(x, y, z, points, showPoints=TRUE, ...) {
#   panel.wireframe(x, y, z, ...)
#   if (showPoints){
#     panel.cloud(points$x, points$y, points$z, ...)
#   }
# }
# wireframe(z ~ x * y, data=myGrid, xlab="X", ylab="Y", zlab="Z",
#           panel=mypanel, points=point, showPoints=FALSE)

# x <- seq(-.8, .8, .1)
# y <- seq(-.8, .8, .1)
# myGrid <- data.frame(expand.grid(x,y))
# colnames(myGrid) <- c("x","y")
# myGrid$z <- myGrid$x + myGrid$y
# noise <- rnorm(length(myGrid$z),.3,.2)
# myGrid$z2 <- myGrid$x + myGrid$y + noise
# mypanel <- function(x,y,z,z2,...) {
#   panel.wireframe(x,y,z,...)
#   panel.cloud(x,y,z2,...)
# }
# wireframe(z ~ x * y, data=myGrid, xlab="X", ylab="Y", zlab="Z",
#           panel=mypanel, z2=myGrid$z2)

@

### 2013-03-30 EconGrowthSeminar.Rnw
Now possible to include the optimum point on the SSE vs. lambda and gamma graph.

\date[Math Seminar] % (optional)
{Mathematics Colloquium\\
4 April 2013
}

@@

\subsection[Economic]{Sources of Economic Data} %+++++++++++++++++++++++++++++++++++++++

@@

\begin{frame}{Economic Definitions} %----------------------------
  \begin{itemize}
    \item{Output ($Y$): total value (\$) of goods produced (GDP)}
    \item{Capital Stock ($K$): market value of productive goods (machines, tools, infrastructure).}
    \item{Labor ($L$): workers in an economy (people or hours)}
  \end{itemize}
\end{frame}

\begin{frame}{Sources of Economic Data} %----------------------------
  \begin{center}
    \scriptsize
    \begin{tabular}{r c c l}
      \multirow{3}{*}{WB GEP} & \multirow{3}{*}{\includegraphics[scale=0.07]{figure_other/Figure_World_Bank.png}} & $Y$, $K$ for all countries & \multirow{3}{*}{Real 2005\$} \\
                              &                          & (1980-2011: Developed; & \\
                              &                          & 1991-2011: Other) & \\
      \hline
      \multirow{4}{*}{WB WDI} & \multirow{4}{*}{\includegraphics[scale=0.07]{figure_other/Figure_World_Bank.png}} & Pop, Pop < 15 yo,    & \multirow{4}{*}{Total Employment} \\
                              &                          & Emp/Pop ratio > 15 yo, &                                \\
                              &                          & used to find $L$, (1991-2011: &                         \\
                              &                          & CN, ZA, SA, IR, TZ, ZM) &                         \\
      \hline
      \multirow{2}{*}{US BLS} & \multirow{2}{*}{\includegraphics[scale=0.07]{figure_other/Figure_BLS.png}} & $L$ for developed countries & Labor hours \\
                              &                          & (1980-2011: US, UK, JP)     & per year \\ 
    \end{tabular}
    \normalsize
  \end{center}
\end{frame}

\subsection[Energy]{Sources of Energy Data} %+++++++++++++++++++++++++++++++++++++++

@@

\begin{frame}{Energy Definitions} %----------------------------
  \begin{itemize}
    \item{Thermal energy ($Q$): heat content of an energy carrier}
    \item{Exergy ($X$): maximum work from a heat carrier}
    \item{Useful work ($U$): energy that generates economic activity directly}
  \end{itemize}
  \begin{center}Note: indexed $Q$ and $X$ are identical unless energy mix changes significantly. \end{center}
\end{frame}

@@

\subsection[Historical Data]{Historical Data} %+++++++++++++++++++++++++++++++++++++++

\minorSeparator{Historical Data}

@@

    \begin{itemize}
      \item{$\mathrm{e}^{\lambda(t-t_0)}$: Time-dependent technology term}
      \begin{itemize}
        \item{Represents the irrepressible ``progress'' of technology}
        \item{Is independent of factor of production ($f$)}
      \end{itemize}
      \item{Used for illustration purposes here}
      \item{Example: \citet{Nel:2009wq} use $f$ = $q$}
      \item{$\lambda$ and $m$ estimated by \texttt{nls} function in \texttt{R}}
    \end{itemize}
\end{frame}

\begin{frame}{Curve-fitting as Optimization} %----------------------------
  \begin{itemize}
    \item{Select values of parameters to minimize errors ($SSE$) between predicted and historical economic output ($y$)}
    \vspace{\baselineskip}

    \item{Example: Single-factor Model with $Q$}
    \begin{itemize}
      \item{$y = \mathrm{e}^{\lambda (t-t_0)}q^{\gamma}$}
      \item{Select $\lambda$ and $\gamma$ to minimize\\
            $SSE = \sum\limits_{t=1980}^{2011}(y_{pred} - y_{act})^2$}
    \end{itemize}
    \vspace{\baselineskip}

    \item{``Curve-fitting,'' ``parameter estimation,''  ``optimization,'' and ``minimization'' are used interchangeably}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Comments on Single-factor Model with $Q$} %----------------------------
  \begin{itemize}
    \item{Not typically used for economic modeling. Illustrates confidence intervals.}
    \vspace{\baselineskip}

@@

\subsection[Cobb-Douglas]{Evaluation of Cobb-Douglas} %+++++++++++++++++++++++++++++++++++++++

@@

\subsection[CES]{Evaluation of CES} %+++++++++++++++++++++++++++++++++++++++

@@

\begin{frame}{Comments on CES Model Equation} %----------------------------
  \begin{center} \small \cesEqn \normalsize \end{center}
  \begin{itemize}
    \item{$\mathrm{e}^{\lambda(t-t_0)}$: Time-dependent technology term}
    \item{$\delta$: split between ($kl$) and ($e$)}
    \item{$\sigma \equiv \frac{1}{1+\rho}$: elasticity of substitution ($0 \leq \sigma \leq 1$)}
    \begin{itemize}
      \item{$\sigma \rightarrow 0$: ($kl$) and ($e$) are complementary}
      \item{$\sigma \rightarrow 1$: ($kl$) and ($e$) are substitutes}
    \end{itemize}
    \item{Example: \citet{Stern:2012ey}}
    \item{$\gamma$, $\lambda$, $\delta_1$, $\rho_1$, $\delta$, $\rho$ estimated by \texttt{cesEst} function in \texttt{R}'s \texttt{micEcon} package}
  \end{itemize}
\end{frame}

@@

\subsection[LINEX]{Evaluation of LINEX} %+++++++++++++++++++++++++++++++++++++++

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Summary}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

>>>>

\date[Math Seminar] % (optional)
{Mathematics Colloquium\\
Thursday, 4 April 2013
}

@@

\subsection[Economic]{Economic} %+++++++++++++++++++++++++++++++++++++++

@@

\begin{frame}{Economic Definitions} %----------------------------
  \begin{itemize}
    \item{Output ($Y$): total value (\$) of goods produced (GDP)}
    \vspace{\baselineskip}

    \item{Capital stock ($K$): market value of productive goods (e.g., machines, tools, infrastructure)}
    \vspace{\baselineskip}

    \item{Labor ($L$): workers in an economy (people or hours)}
  \end{itemize}
\end{frame}

\begin{frame}{Sources of Economic Data} %----------------------------
  \begin{center}
    \scriptsize
    \begin{tabular}{r c c l}
      \multirow{3}{*}{WB GEP} & \multirow{3}{*}{\includegraphics[scale=0.07]{figure_other/Figure_World_Bank.png}} & $Y$, $K$ for all countries & \multirow{3}{*}{Real 2005\$} \\
                              &                          & (1980-2011: Developed; & \\
                              &                          & 1991-2011: Other) & \\
      \hline
      \multirow{4}{*}{WB WDI} & \multirow{4}{*}{\includegraphics[scale=0.07]{figure_other/Figure_World_Bank.png}} & Pop, Pop < 15 yo,    & \multirow{4}{*}{Total employment} \\
                              &                          & Emp/Pop ratio > 15 yo, &                                \\
                              &                          & used to find $L$, (1991-2011: &                         \\
                              &                          & CN, ZA, SA, IR, TZ, ZM) &                         \\
      \hline
      \multirow{2}{*}{US BLS} & \multirow{2}{*}{\includegraphics[scale=0.07]{figure_other/Figure_BLS.png}} & $L$ for developed countries & Labor hours \\
                              &                          & (1980-2011: US, UK, JP)     & per year \\ 
    \end{tabular}
    \normalsize
  \end{center}
\end{frame}

\subsection[Energy]{Energy} %+++++++++++++++++++++++++++++++++++++++

@@

\begin{frame}{Energy Definitions} %----------------------------
  \begin{itemize}
    \item{Thermal energy ($Q$): heat content of an energy carrier}
    \vspace{\baselineskip}

    \item{Exergy ($X$): maximum work from a heat carrier}
    \vspace{\baselineskip}

    \item{Useful work ($U$): energy that generates economic activity directly}
  \end{itemize}
  \begin{center}Note: indexed $Q$ and $X$ are identical unless energy mix changes significantly \end{center}
\end{frame}

@@

\subsection[Visualization]{Visualization} %+++++++++++++++++++++++++++++++++++++++

\minorSeparator{Visualization of Historical Data}

@@

    \begin{itemize}
      \item{Illustrative}
      \item{$\mathrm{e}^{\lambda(t-t_0)}$: Time-dependent technology term}
      \begin{itemize}
        \item{Irrepressible ``progress'' of technology}
        \item{Independent of factor of production ($f$)}
      \end{itemize}
      \item{Example: \citet{Nel:2009wq} use $f$ = $q$}
      \item{$\lambda$ and $m$ estimated by \texttt{nls} function in \texttt{R}}
    \end{itemize}
\end{frame}

\begin{frame}{Example: Single-factor model with $q$} %----------------------------
  \begin{center}$y = \mathrm{e}^{\lambda (t-t_0)}q^{\gamma}$\end{center}
  \begin{itemize}
    \item{Select $\lambda$, $\gamma$}
    \item{To minimize \\ $SSE = \sum\limits_{t=1980}^{2011}(y_{pred} - y_{act})^2$}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Comments on Single-factor Model with $Q$} %----------------------------
  \begin{itemize}
    \item{Not typically used for economic modeling: illustrates confidence intervals}
    \vspace{\baselineskip}

@@

\subsection[Cobb-Douglas]{Cobb-Douglas} %+++++++++++++++++++++++++++++++++++++++

@@

\subsection[CES]{CES} %+++++++++++++++++++++++++++++++++++++++

@@

\begin{frame}{Comments on CES Model Equation} %----------------------------
  \begin{center} \small \cesEqn \normalsize \end{center}
  \begin{itemize}
    \item{$\mathrm{e}^{\lambda(t-t_0)}$: Time-dependent technology term}
    \item{$\delta$: split between ($kl$) and ($e$)}
    \item{Example: \citet{Stern:2012ey}}
    \item{$\gamma$, $\lambda$, $\delta_1$, $\rho_1$, $\delta$, $\rho$ estimated by \texttt{cesEst} function in \texttt{R}'s \texttt{micEcon} package}
  \end{itemize}
\end{frame}

@@

\subsection[LINEX]{LINEX} %+++++++++++++++++++++++++++++++++++++++

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Summary}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\minorSeparator{CES Model}

\begin{frame}{CES Elasticity of Substitution} %----------------------------
  \begin{center} \small \cesEqn \normalsize \end{center}
  \begin{itemize}
    \item{$\sigma \equiv \frac{1}{1+\rho}$: elasticity of substitution ($0 \leq \sigma \leq 1$)}
    \begin{itemize}
      \item{$\sigma \rightarrow 0$: ($kl$) and ($e$) are complementary}
      \item{$\sigma \rightarrow 1$: ($kl$) and ($e$) are substitutes}
    \end{itemize}
  \end{itemize}
\end{frame}
